Discovering the Rules of language using Bayesian Program Learning

### 2018-06-16 outline.tex
science paper progress

@@

\pagebreak
These problems are solved:
\begin{itemize}
\item Kikurai: fricatives alternate with stops after nasals
\item Greek: Velar stops alternate with palletized versions before front vowels
\item Farsi: Trills alternate with flaps
\item Osage: coronal stops become dentals before central vowels
\item Amharic: alternation between \textipa{@} \& \textipa{E}
\item Gen: l/r alternation
\item Kishambaa: voiced/unvoiced nasals alternate
\item  Thai: stops are unreleased word finally
\item Palauan: a word initial neutralizing rule
\item Quechua: Velar becomes uvular when followed by uvular (spreading)
\item Lhasa Tibetan: no contrast between velar/uvular, or voiced/voiceless stops or fricatives
\item Axininca Campa: stops become glides
\item Kikuyu: infinitive prefix can surface as either k or \textipa{G}
\item Korean: vowel harmony, aspiration only surfaces in certain contexts
\item Hungarian: vowel harmony and voicing assimilation
\item Kikuria: vowel harmony
\item Farsi: a deletion rule explains singular/plural
\item Tibetan: initial consonant cluster reduction explains counting system
\item Makonde: stress patterns; unstressed vowels are neutralized
\item North Saami: 3 neutralization rules explain nominative sg essive
\item Samoan: 2 deletion rules that explain words that sound the same in one inflection being different in another
\item     Russian: devoicing of word final obstruent
\item English: verbal inflections (voicing assimilation, epenthesis)
\item Finnish: nominative/partive explained by vowel raising and vowel harmony
\item Kerewe: Interacting tone rules
\item Polish: vowel alternations interacting with devoicing of word final obstruent in singular/plural
\item Ancient Greek: voicing assimilation, deaspiration, deletion rule. Order matters.
\item Serbo-Croatian: predictable stress, devoicing, neutralizing, epenthesis
\end{itemize}
17 textbook problems are not solved. 2 of these need a tier-based representation. A couple more of these are due to limitations in the program learner's representation of rules. The rest are because they involve scaling to more than 3 rules, which is about the practical limit for exact search with these methods.
\pagebreak

>>>>

\subsubsection{Bounding the size of the rules}
Sketch only works with finite program spaces, because it unrolls anything that loops or is recursive up to some depth bound. A necessary consequence of this is that we cannot learn rules with an arbitrary number of feature matrices. I bound the number of feature matrices to be 2 on each side.

\subsubsection{``Freezing'' rules and underlying forms}
\textbf{incrementalCEGIS} constructs a sequence of models that explain
more and more of the data.  If in this sequence we see that a certain
rule always occurs, or that a certain surface form always has a
certain underlying representation, then we might guess that rule or
underlying representation is really part of the correct solution and
that we shouldn't go back and try to revise that rule or underlying
form.  Concretely what this means is that I keep track of how many
times a rule or underlying form is found by \textbf{incrementalCEGIS};
once it has occurred at least 5 times I ``freeze'' the rule or
underlying form and do not allow it to change in the future.

For example, after a few rounds of \textbf{incrementalCEGIS}
we might find that we always need to devoice word final segments.
If this is the case then we will stop trying to change the rule that devoices
word final segments.

@@

\pagebreak
These problems are solved:
\begin{itemize}
\item Kikurai: fricatives alternate with stops after nasals
\item Greek: Velar stops alternate with palletized versions before front vowels
\item Farsi: Trills alternate with flaps
\item Osage: coronal stops become dentals before central vowels
\item Amharic: alternation between \textipa{@} \& \textipa{E}
\item Gen: l/r alternation
\item Kishambaa: voiced/unvoiced nasals alternate
\item  Thai: stops are unreleased word finally
\item Palauan: a word initial neutralizing rule
\item Quechua: Velar becomes uvular when followed by uvular (spreading)
\item Lhasa Tibetan: no contrast between velar/uvular, or voiced/voiceless stops or fricatives
\item Axininca Campa: stops become glides
\item Kikuyu: infinitive prefix can surface as either k or \textipa{G}
\item Korean: vowel harmony, aspiration only surfaces in certain contexts
\item Hungarian: vowel harmony and voicing assimilation
\item Kikuria: vowel harmony
\item Farsi: a deletion rule explains singular/plural
\item Tibetan: initial consonant cluster reduction explains counting system
\item Makonde: stress patterns; unstressed vowels are neutralized
\item North Saami: 3 neutralization rules explain nominative sg essive
\item Samoan: 2 deletion rules that explain words that sound the same in one inflection being different in another
\item     Russian: devoicing of word final obstruent
\item English: verbal inflections (voicing assimilation, epenthesis)
\item Finnish: nominative/partive explained by vowel raising and vowel harmony
\item Kerewe: Interacting tone rules
\item Polish: vowel alternations interacting with devoicing of word final obstruent in singular/plural
\item Ancient Greek: voicing assimilation, deaspiration, deletion rule. Order matters.
\item Serbo-Croatian: predictable stress, devoicing, neutralizing, epenthesis
\item Catalan: Spritization, devoicing, several different kinds of cluster reduction
\item Korean: place assimilation and nasalization
\item Latin: assimilation, devoicing, different kinds of cluster reduction, vowel harmony
\end{itemize}
These problems it solves most of:
\begin{itemize}
\item Somali:  Spritization, copying rules
\end{itemize}
12 textbook problems are not solved, along with two other problems that involve tier--based representations.
3 of the above problems are not actually exercises but are instead examples that I included because they are neat problems. This means that we solve $\frac{28}{40} = 70\%$ of the textbook currently,
using a slightly liberal interpretation of what it means to ``solve a problem'' (e.g., our solution for Latin is slightly different from the textbook answer, but to me it looks fine and is at least as compressive).
\pagebreak

### 2018-06-16 test.tex
science paper progress

\documentclass{article}
\usepackage[margin = 0.1cm]{geometry}
\usepackage{tipa}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{longtable}
\begin{document}

@@

\verb|[+vowel]| $\to$ \verb|[+highTone]| / \verb|[+highTone]| \verb|[ ]|* \verb|_| \\$\varnothing$ $\to$ \textipa{a} / \verb|[-vowel]| \verb|_| \verb|[-vowel]| \#\\\verb|[+vowel]| $\to$ \verb|[-highTone]| /  \verb|_| \verb|[ ]|* \verb|[+highTone]|\\\textipa{l} $\to$ \textipa{o} /  \verb|_| \#\\\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|

@@

\verb|[ ]| $\to$ $-2$ / \verb|[ ]| \verb|[ ]| \verb|_| \verb|[+sibilant]| \#\\\verb|[ ]| $\to$ \textipa{e} / \verb|[+palletized]| \verb|_| \verb|[+voice]| \verb|[+vowel]|\\\verb|[-glide]| $\to$ \verb|[-palletized]| /  \verb|_| \verb|[-back]|\\\textipa{w} $\to$ \textipa{v} / \verb|[+vowel]| \verb|_| \verb|[+vowel]|\\\verb|[ ]| $\to$ \verb|[+palletized]| /  \verb|_| \textipa{i}\\\textipa{o} $\to$ \textipa{i} /  \verb|_| \verb|[-vowel]|* \verb|[-vowel]| \#\\\textipa{l} $\to$ \textipa{w} /  \verb|_| \#

@@

\emph{Latin:}\\\begin{longtable}{lll|l}\toprule
$\varnothing$ $+$stem$+$ \textipa{s} & $\varnothing$ $+$stem$+$ \textipa{is} & $\varnothing$ $+$stem$+$ \textipa{i:} & UR
\\ \midrule
\textipa{arks} & \textipa{arkis} &  --  &  -- \\
\textipa{duks} & \textipa{dukis} &  --  &  -- \\
\textipa{daps} & \textipa{dapis} &  --  &  -- \\
\textipa{re:ks} & \textipa{re:gis} &  --  &  -- \\
\textipa{falanks} & \textipa{falangis} &  --  &  -- \\
\textipa{filiks} & \textipa{filikis} &  --  &  -- \\
\textipa{lapis} & \textipa{lapidis} &  --  &  -- \\
\textipa{li:s} & \textipa{li:tis} &  --  &  -- \\
\textipa{fraws} & \textipa{frawdis} &  --  &  -- \\
\textipa{noks} & \textipa{noktis} &  --  &  -- \\
\textipa{frons} & \textipa{frontis} &  --  &  -- \\
\textipa{frons} & \textipa{frondis} &  --  &  -- \\
\textipa{inku:s} & \textipa{inku:dis} &  --  &  -- \\
\textipa{sors} & \textipa{sortis} &  --  &  -- \\
\textipa{fu:r} & \textipa{fu:ris} &  --  &  -- \\
\textipa{murmur} & \textipa{murmuris} &  --  &  -- \\
\textipa{augur} & \textipa{auguris} &  --  &  -- \\
\textipa{arbor} & \textipa{arboris} &  --  &  -- \\
\textipa{pugil} & \textipa{pugilis} &  --  &  -- \\
\textipa{sal} & \textipa{salis} &  --  &  -- \\
\textipa{adeps} & \textipa{adipis} &  --  &  -- \\
\textipa{apeks} & \textipa{apikis} &  --  &  -- \\
\textipa{pri:nkeps} & \textipa{pri:nkipis} &  --  &  -- \\
\textipa{ekwes} & \textipa{ekwitis} &  --  &  -- \\
\textipa{miles} & \textipa{militis} &  --  &  -- \\
\textipa{no:men} & \textipa{no:minis} &  --  &  -- \\
\textipa{karmen} & \textipa{karminis} &  --  &  -- \\
\textipa{lu:men} & \textipa{lu:minis} &  --  &  -- \\
\textipa{wenter} & \textipa{wentris} &  --  &  -- \\
\textipa{pater} & \textipa{patris} &  --  &  -- \\
\textipa{kada:wer} & \textipa{kada:weris} &  --  &  -- \\
\textipa{tu:ber} & \textipa{tu:beris} &  --  &  -- \\
\textipa{piper} & \textipa{piperis} &  --  &  -- \\
\textipa{karker} & \textipa{karkeris} &  --  &  -- \\
\textipa{die:s} &  --  & \textipa{die:i:} &  -- \\
\textipa{li:ber} &  --  & \textipa{li:beri:} &  -- \\
\textipa{miser} &  --  & \textipa{miseri:} &  -- \\
\textipa{ager} &  --  & \textipa{agri:} &  -- \\
\textipa{sinister} &  --  & \textipa{sinistri:} &  -- \\
\textipa{liber} &  --  & \textipa{libri:} &  -- \\
\textipa{as} & \textipa{assis} &  --  &  -- \\
\textipa{os} & \textipa{ossis} &  --  &  -- \\
\textipa{far} & \textipa{farris} &  --  &  -- \\
\textipa{mel} & \textipa{mellis} &  --  &  -- \\
\textipa{o:s} & \textipa{o:ris} &  --  &  -- \\
\textipa{flo:s} & \textipa{flo:ris} &  --  &  -- \\
\textipa{mu:s} & \textipa{mu:ris} &  --  &  -- \\
\textipa{cru:s} & \textipa{cru:ris} &  --  &  -- \\
\textipa{kinis} & \textipa{kineris} &  --  &  -- \\
\textipa{pulvis} & \textipa{pulveris} &  --  &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|\\\textipa{e} $\to$ $2$ /  \verb|_| \verb|[-liquid]| \verb|[+vowel]|\\\verb|[-vowel]| $\to$ $\varnothing$ / \verb|[+coronal]| \verb|_| \#\\\verb|[-sonorant]| $\to$ \textipa{s} /  \verb|_| \#\\$\varnothing$ $\to$ \textipa{e} / \verb|[-sonorant]| \verb|_| \verb|[+voice]| \#\\$\varnothing$ $\to$ $2$ / \verb|[+sibilant]| \verb|_| \verb|[+vowel]| \verb|[-voice]|\\$\varnothing$ $\to$ $1$ / \# \verb|[-coronal]| \verb|[-high]| \verb|_| \verb|[+voice]| \textipa{i}
\end{tabular}

>>>>

\documentclass{article}
\usepackage[margin = 0.1cm]{geometry}
\usepackage{phonrule}
\usepackage{tipa}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{longtable}
\begin{document}

@@

\phonl{V}{[ +highTone ]}{[ +highTone ] [ ]\textsubscript{*}}\\\phonb{$\varnothing$}{\textipa{a}}{C}{C \#}\\\phonr{V}{[ -highTone ]}{[ ]\textsubscript{*} [ +highTone ]}\\\phonr{\textipa{l}}{\textipa{o}}{\#}\\\phonr{[ -sonorant ]}{[ -voice ]}{[ -voice ]}
\begin{tabular}{l}\emph{Textbook solution rules: }\\
\phonl{V}{[ +highTone ]}{[ +highTone ] C\textsubscript{*}}\\\phonr{V}{[ -highTone ]}{C\textsubscript{*} [ +highTone ]}\\\phonb{$\varnothing$}{\textipa{a}}{C}{C \#}\\\phonr{\textipa{l}}{\textipa{o}}{\#}\\\phonr{[ -sonorant ]}{[ -voice ]}{[ -voice ]}

@@

\phonb{[ ]}{[ ]\textsubscript{i}}{[ ]\textsubscript{i} [ ]}{[ +sibilant ] \#}\\\phonb{[ ]}{\textipa{e}}{[ +palletized ]}{[ +voice ] V}\\\phonr{[ -glide ]}{[ -palletized ]}{[ -back ]}\\\phonb{\textipa{w}}{\textipa{v}}{V}{V}\\\phonr{[ ]}{[ +palletized ]}{\textipa{i}}\\\phonr{\textipa{o}}{\textipa{i}}{C\textsubscript{*} C \#}\\\phonr{\textipa{l}}{\textipa{w}}{\#}
\begin{tabular}{l}\emph{Textbook solution rules: }\\
\phonr{C}{[ +palletized ]}{\textipa{i}}\\\phonl{\textipa{o}}{\textipa{e}}{[ +palletized ] $+$}\\\phonr{[ +vowel -low -high ]}{\textipa{i}}{C\textsubscript{*} C \#}\\\phonr{[ -glide -vowel ]}{[ -palletized ]}{\textipa{e}}\\\phonr{\textipa{l}}{\textipa{w}}{\#}\\\phonr{\textipa{v}}{\textipa{w}}{\#}

@@

\emph{Latin:}\\\begin{longtable}{lll|l}\toprule
$\varnothing$ $+$stem$+$ \textipa{s} & $\varnothing$ $+$stem$+$ \textipa{is} & $\varnothing$ $+$stem$+$ \textipa{i:} & UR
\\ \midrule
\textipa{arks} & \textipa{arkis} &  --  &  -- \\
\textipa{duks} & \textipa{dukis} &  --  &  -- \\
\textipa{daps} & \textipa{dapis} &  --  &  -- \\
\textipa{re:ks} & \textipa{re:gis} &  --  &  -- \\
\textipa{falanks} & \textipa{falangis} &  --  &  -- \\
\textipa{filiks} & \textipa{filikis} &  --  &  -- \\
\textipa{lapis} & \textipa{lapidis} &  --  &  -- \\
\textipa{li:s} & \textipa{li:tis} &  --  &  -- \\
\textipa{fraws} & \textipa{frawdis} &  --  &  -- \\
\textipa{noks} & \textipa{noktis} &  --  &  -- \\
\textipa{frons} & \textipa{frontis} &  --  &  -- \\
\textipa{frons} & \textipa{frondis} &  --  &  -- \\
\textipa{inku:s} & \textipa{inku:dis} &  --  &  -- \\
\textipa{sors} & \textipa{sortis} &  --  &  -- \\
\textipa{fu:r} & \textipa{fu:ris} &  --  &  -- \\
\textipa{murmur} & \textipa{murmuris} &  --  &  -- \\
\textipa{augur} & \textipa{auguris} &  --  &  -- \\
\textipa{arbor} & \textipa{arboris} &  --  &  -- \\
\textipa{pugil} & \textipa{pugilis} &  --  &  -- \\
\textipa{sal} & \textipa{salis} &  --  &  -- \\
\textipa{adeps} & \textipa{adipis} &  --  &  -- \\
\textipa{apeks} & \textipa{apikis} &  --  &  -- \\
\textipa{pri:nkeps} & \textipa{pri:nkipis} &  --  &  -- \\
\textipa{ekwes} & \textipa{ekwitis} &  --  &  -- \\
\textipa{miles} & \textipa{militis} &  --  &  -- \\
\textipa{no:men} & \textipa{no:minis} &  --  &  -- \\
\textipa{karmen} & \textipa{karminis} &  --  &  -- \\
\textipa{lu:men} & \textipa{lu:minis} &  --  &  -- \\
\textipa{wenter} & \textipa{wentris} &  --  &  -- \\
\textipa{pater} & \textipa{patris} &  --  &  -- \\
\textipa{kada:wer} & \textipa{kada:weris} &  --  &  -- \\
\textipa{tu:ber} & \textipa{tu:beris} &  --  &  -- \\
\textipa{piper} & \textipa{piperis} &  --  &  -- \\
\textipa{karker} & \textipa{karkeris} &  --  &  -- \\
\textipa{die:s} &  --  & \textipa{die:i:} &  -- \\
\textipa{li:ber} &  --  & \textipa{li:beri:} &  -- \\
\textipa{miser} &  --  & \textipa{miseri:} &  -- \\
\textipa{ager} &  --  & \textipa{agri:} &  -- \\
\textipa{sinister} &  --  & \textipa{sinistri:} &  -- \\
\textipa{liber} &  --  & \textipa{libri:} &  -- \\
\textipa{as} & \textipa{assis} &  --  &  -- \\
\textipa{os} & \textipa{ossis} &  --  &  -- \\
\textipa{far} & \textipa{farris} &  --  &  -- \\
\textipa{mel} & \textipa{mellis} &  --  &  -- \\
\textipa{o:s} & \textipa{o:ris} &  --  &  -- \\
\textipa{flo:s} & \textipa{flo:ris} &  --  &  -- \\
\textipa{mu:s} & \textipa{mu:ris} &  --  &  -- \\
\textipa{cru:s} & \textipa{cru:ris} &  --  &  -- \\
\textipa{kinis} & \textipa{kineris} &  --  &  -- \\
\textipa{pulvis} & \textipa{pulveris} &  --  &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\phonr{[ -sonorant ]}{[ -voice ]}{[ -voice ]}\\\phonr{\textipa{e}}{V\textsubscript{i}}{[ -liquid ] V\textsubscript{i}}\\\phonb{C}{$\varnothing$}{[ +coronal ]}{\#}\\\phonr{[ -sonorant ]}{\textipa{s}}{\#}\\\phonb{$\varnothing$}{\textipa{e}}{[ -sonorant ]}{[ +voice ] \#}\\\phonb{$\varnothing$}{[ -voice ]\textsubscript{i}}{[ +sibilant ]}{V [ -voice ]\textsubscript{i}}\\\phonb{$\varnothing$}{[ +voice ]\textsubscript{i}}{\# [ -coronal ] [ -high ]}{[ +voice ]\textsubscript{i} \textipa{i}}
\end{tabular}
\begin{tabular}{l}\emph{Textbook solution rules: }\\
\phonr{[ -sonorant ]}{[ -voice ]}{[ -voice ]}\\\phonl{\textipa{s}}{$\varnothing$}{[ +coronal +sonorant ]}\\\phonr{[ +coronal -sonorant ]}{$\varnothing$}{\textipa{s}}\\\phonr{\textipa{e}}{\textipa{i}}{C V}\\\phonb{$\varnothing$}{\textipa{e}}{C}{[ +coronal +sonorant ] \#}\\\phonb{\textipa{s}}{\textipa{r}}{V}{V}\\\phonr{[ +vowel -back ]}{[ -high ]}{\textipa{r}}\\\phonr{$1$}{$\varnothing$}{C \#}
\end{tabular}
